Ignition of Fuel Beds by Firebrands

### 2022-04-08 manuscript3.tex
Updates to species paper results

\section{Methods}
\label{sec:methods3}
    The 50\% probability of flaming ignition with respect to the surface temperature of a resistance heater was determined for five different materials. The fuel bed materials used were Douglas-fir wood, pine wood, Douglas-fir bark, wheat straw, and oak wood. These materials were chosen to represent a range of chemical compositions, and are materials that may be subject to ember storms in a wildland urban interface fire. Table \ref{tab:composition} shows the estimated chemical composition of the materials based on the Bioenergy Feedstock Library~\cite{feedstock}. To create fuel beds of consistent particle size,  materials were granulated and then sorted such that the particles fit through a screen with openings of 2.1\si{\milli\meter} but not through openings of 0.85\si{\milli\meter}. Once sorted, the materials were oven dried at 103\si{\celsius}.
    \begin{table*}[hpbt]
        \caption{Proportion of cellulose, hemicellulose, and lignin of the materials tested estimated from the Bioenergy Feedstock Library~\cite{feedstock}}
        \centering
        \begin{tabular}{crrrr}
            % \rowcolor{gray!50}
            Material & Cellulose (\%) & Hemicellulose (\%) & Lignin (\%) & Other (\%) \\
            \hline
            Oak Wood         & 45 & 22 & 24 & 9\\
            Douglas-fir Wood & 44 & 22 & 28 & 6\\
            Wheat Straw      & 44 & 27 & 19 & 10\\
            Pine Wood        & 40 & 30 & 20 & 10\\
            Douglas-Fir Bark & 26 & 11 & 59 & 4
        \end{tabular}
        \label{tab:composition}
    \end{table*}

    The experimental apparatus, shown in Figure~\ref{fig:apparatus} was designed to replicate an ember landing on a fuel bed. A lever arm holds the cartridge heater in a fixed position above the fuel bed for the duration of the test. The ember, represented by a 6.35\si{\milli\meter} diameter 51\si{\milli\meter} long cartridge heater capable of a 250\si{\watt} output, was inserted 3\si{\milli\meter} into the bed (approximately half the diameter of the heater). The temperature of the heater was recorded with a type-K thermocouple for the duration of each test. 
    The temperature of the heater ranged from 250\si{\celsius} to 750\si{\celsius} and was controlled using PID control logic implemented in LabVIEW. The PID controller kept the heater temperature within 6\% of the setpoint for the duration of the tests. The temperature setpoints of the resistance heater were determined using the three-phase optimal design procedure to efficiently determine the desired ignition probability for the number of tests conducted~\cite{Wu2014Three-phaseExperiments}. For tests where flaming ignition did not occur data was recorded for 3000\si{\second} or until the reaction front of the smoldering material reached the edge of the test container. The fixed position of the cartridge heater and the use of the cartridge heater was implemented to maintain consistency between tests of the same material and between test series of different materials. The use of the cartridge heater and the fixed position is not necessarily representative of a burning ember landing on the fuel bed but the consistency of surface temperatures, ability to record temperature, and consistent contact with the fuel bed are essential for comparing results across test series and materials. Flaming ignition was detected with the use of a BPX 65 photodiode sampled at 1 kHz. If the intensity detected rose above a set threshold flaming ignition was said to occur. Visual detection of flames was also used in cohort with the photodiode measurements. The 50\% probability of ignition was determined from a logistic regression on the outcomes of 25 ignition tests conducted for each of the materials for which flaming ignition was observed. The scikit-learn python package~\cite{scikit-learn} was used to perform the logistic regressions used to predict ignition probabilities.

@@

        \label{fig:apparatus}

@@

\section{Results and Discussion}
    \label{sec:results3}
    The flaming or non-flaming ignition outcome of each test and the logistic regression results for Douglas-fir wood, oak wood, and pine wood are shown in Figure~\ref{fig:logistic_plot}. The circular markers represent the outcome of each test, either flaming or no ignition. The logistic regression for each material is shown as a solid line with the shaded regions representing the 95\% confidence interval for each regression. Douglas-fir bark and wheat straw are not shown in Figure~\ref{fig:logistic_plot} because flaming ignition was not observed for five tests at the maximum heater temperature of 750\si{\celsius}. The heater temperature estimated to produce 50\% ignition probability from the logistic regressions is 574\si{\celsius} for the Douglas-fir wood, 634\si{\celsius} for the pine wood, and 662\si{\celsius} for the oak wood. Anecdotally, self sustained smoldering was observed for all five materials at temperatures lower than the flaming ignition temperature. The heater temperature corresponding the onset of smoldering ignition was not measured in this study. 
        \begin{figure}[htpb]
            \centering
            \includegraphics[width=\figureWidthSet, trim={0 0 37cm 0}, clip]{conference_results_binary.png}
            \caption{Test results for materials where flaming ignition was observed. The circular markers denote individual tests. The solid lines represent the logistic regression and the shaded zones the 95\% confidence interval with the 50\% probability of ignition labelled}
            \label{fig:logistic_plot}
        \end{figure}
    Two aspects of the ignition results are of note. As anticipated, the differences in estimated heater temperature to produce 50\% ignition probability suggest that significant differences in ignition occur across a range of materials in the same apparatus and experimental conditions. Second, the transition between no ignition and flaming ignition may become less defined (i.e., the 95\% confidence interval of the regression spans a much larger temperature range) as the ignition temperature decreases. This is significant because it may cause more difficulty in predicting the ignition of materials that ignite at lower temperatures. Lower confidence in predicting ignition for easily ignitable materials would be detrimental to the usefulness of a model or predictive tool that may be implemented from this testing methodology. It is, however, unclear if the ignition to no ignition transition in other materials of similarly low ignition temperature will behave similarly to that of Douglas-fir. Characterization of additional materials is needed to elucidate these trends.

    Figure~\ref{fig:composition_plot} shows the averaged cellulose, hemicellulose, and lignin concentrations for each material according to the samples recorded in the Bioenergy Feedstock Library as shown in Table~\ref{tab:composition}. The marker colors for each material correspond to the calculated 50\% ignition probability except for the wheat straw and Douglas-fir bark which are denoted as 800\si{\celsius} to indicate that the 50\% ignition probability was higher than the temperatures tested.  Note that due to scaling inherent to a ternary projection the proportion of the fuels characterized as other causes a shift in the data. For example, the wheat straw and Douglas-fir are estimated to have the same cellulose content but do not lie on the same iso-line of 44\% cellulose. Nonetheless, the relative proportions of each constituent are retained and comparisons may still be made. 

@@

    Perhaps unsurprisingly, Douglas-fir bark, with the highest lignin content, was not observed to ignite for the conditions tested. This suggests that materials with high lignin contents may be the least likely to ignite when exposed to embers and thus are the safest with respect to flaming ignition risk around homes. This does not, however, account for the potential for smoldering ignition and a subsequent smoldering to flaming transition that may occur. Thermogravametric analysis experiments have shown that while lignin begins to decompose at the lowest temperature of the three primary components, it decomposes at the slowest rate and at a much wider range of temperatures~\cite{Yang2007}. It was also observed that the peak gas production of \ce{CO}, \ce{CO2}, and \ce{CH4} occurs at a higher temperatures than hemicellulose and cellulose. In contrast to Douglas-fir bark, wheat straw had the lowest estimated lignin concentration but was also not observed to ignite. When comparing wheat straw to Douglas-fir wood the higher lignin concentration suggests that Douglas-fir wood is likely to have a higher ignition temperature than wheat straw. This discrepancy is attributed to two material attributes. First, the pyrolysis process is complex and there is not a clear linear relationship between the composition of the fuel and flaming ignition. It appears that materials of different compositions are capable of producing gaseous products of near equal ignitability but the proportion of each constituent for which ignition to occurs most readily is unclear. Second, the thermal conductivity of the materials likely varies significantly which impacts the temperature gradients and mass of the material above the pyrolysis temperature, further obfuscating the effect of composition on ignition. Fortunately, both thermal conductivity and chemical composition are measurable material properties and can be simultaneously considered when studying ignition. In summary, characterizing the flaming ignition propensity of fuels with respect to chemical composition alone will likely not yield satisfactory results. The inclusion of other material properties, such as thermal conductivity, however, may add enough additional knowledge for comprehensive predictions of ignition to be made. Further investigation of the differences in chemical and thermal properties of the materials is necessary to unravel the interactions between thermal and chemical processes and the resulting impact on flaming ignition before knowledge sufficient for a predictive tool that may be used for fire prevention and management may be created

@@

    \begin{table*}[hpbt]
        \caption{Heater temperature required for 50\% probability of ignition for the materials and conditions tested.}
        \centering
        \begin{tabular}{crrr}
            % \rowcolor{gray!50}
            Material & 0.1 \si{\meter\per\second}(\si{\celsius}) & 5.8 \si{\meter\per\second} (\si{\celsius}) & $\Delta$T (\si{\celsius})\\
            \hline
            Douglas-fir Wood & 574  & 391 & -185 \\
            Pine Wood        & 634  & 430 & -204\\
            Oak Wood         & 663  & \textgreater750 & 87\\
            Wheat Straw      & \textgreater750 & \textgreater750 & -\\
            Pine Bark        & \textgreater750 & \textgreater750 & -\\
            Douglas-Fir Bark & \textgreater750 & \textgreater750 & -
        \end{tabular}
        \label{tab:composition}
    \end{table*}

>>>>

\section{Methods}
\label{sec:methods3}
    The 50\% probability of flaming ignition with respect to the surface temperature of a resistance heater was determined for five different materials in quiescent conditions and at a wind speed of 5.8\si{\meter\per\second}. The fuel bed materials used were Douglas-fir wood, Douglas-fir bark, pine wood, pine bark, wheat straw, and oak wood. These materials were chosen to represent a range of chemical compositions, and are materials that may be subject to ember attack in a wildland urban interface fire. Table \ref{tab:composition} shows the estimated chemical composition of the materials based on the Bioenergy Feedstock Library~\cite{feedstock}. To create fuel beds of consistent particle size,  materials were granulated and then sorted such that the particles fit through a screen with openings of 2.1\si{\milli\meter} but not through openings of 0.85\si{\milli\meter}. Once sorted, the materials were oven dried at 103\si{\celsius}.
    \begin{table*}[hpbt]
        \caption{Proportion of cellulose, hemicellulose, and lignin of the materials tested estimated from the Bioenergy Feedstock Library~\cite{feedstock}}
        \begin{tabular}{crrrr}
            % \rowcolor{gray!50}
            Material & Cellulose (\%) & Hemicellulose (\%) & Lignin (\%) & Other (\%) \\
            \hline
            Oak Wood         & 45 & 22 & 24 & 9\\
            Douglas-fir Wood & 44 & 22 & 28 & 6\\
            Wheat Straw      & 44 & 27 & 19 & 10\\
            Pine Wood        & 40 & 30 & 20 & 10\\
            Pine Bark        & 39 & 17 & 37 & 7\\
            Douglas-Fir Bark & 26 & 11 & 59 & 4 
        \end{tabular}
        \label{tab:composition}
    \end{table*}
    Two experimental apparatus were used to conduct ignition experiments. The experimental apparatus used for the tests in quiescent conditions is shown in Figure~\ref{fig:speciesApparatus}. The quiescent apparatus used a lever arm to hold the cartridge heater in a fixed position the duration of the test. The heater was inserted into the fuel bed approximately half of the heater diameter (3\si{\milli\meter}). Tests conducted with a wind speed of \si{\meter\per\second} were performed using the apparatus shown in Figure~\ref{fig:speciesWindTunnel}. This apparatus was operated inside of a wind tunnel where the heater was lowered using an automated lowering device that maintained a constant pressure equivalent to that of a 10\si{\gram} ember landing on the fuel bed. The pressure was monitored and adjusted using a PID control system with force measured by a load cell. For all tests the cartridge heater (firebrand surrogate) used was a 50\si{\milli\meter} long 6.35\si{\milli\meter} diameter cartridge heater.

@@

        \label{fig:speciesApparatus}

@@

    The temperature of the heater was recorded with a type-K thermocouple for the duration of each test. The temperature of the heater for ranged from 250\si{\celsius} to 750\si{\celsius} and was controlled using PID control logic implemented in LabVIEW. The PID controller kept the heater temperature within 6\% of the setpoint for the duration of the tests. The temperature setpoints of the resistance heater were determined using the three-phase optimal design procedure to efficiently determine the desired ignition probability for the number of tests conducted~\cite{Wu2014}. For tests where flaming ignition did not occur data was recorded for 3000\si{\second} or until the reaction front of the smoldering material reached the edge of the test container. The fixed position of the cartridge heater and the use of the cartridge heater was implemented to maintain consistency between tests of the same material and between test series of different materials. The use of the cartridge heater and the fixed position is not necessarily representative of a burning ember landing on the fuel bed but the consistency of surface temperatures, ability to record temperature, and consistent contact with the fuel bed are essential for comparing results across test series and materials. For the quiescent tests flaming ignition was detected with the use of a BPX 65 photodiode sampled at 1 kHz. If the intensity detected rose above a set threshold flaming ignition was said to occur. For the wind tunnel tests flaming ignition was determined from the rising temperature of the cartridge heater in the presence of flame. Visual detection of flames was also used in cohort with the photodiode measurements. The 50\% probability of ignition was determined from a logistic regression on the outcomes of 25 ignition tests conducted for each of the materials for which flaming ignition was observed. The scikit-learn python package~\cite{scikit-learn} was used to perform the logistic regressions used to predict ignition probabilities.
    \begin{figure}[hbpt]
            \resizebox{0.5\columnwidth}{!}{%
                    \filldraw[pattern=north west lines, pattern color=brown, thick] (2.84, 0)  rectangle (4.34, -.75) node[pos=0.5,rectangle,fill=white] {\scriptsize Fuel Bed};
                    \fill [draw=white,  inner color=red, outer color=white ] (3.59, 0.01) circle (0.15);
                    \filldraw[draw=black,fill=white, thick] (0, 0)      rectangle (6, 3);
                    \fill[fill=black!50] (3.58, 0) rectangle (3.61, 1.52);
                    \fill[fill=black] (3.55, 1.52) rectangle (4.23, 1.57);
                    \filldraw[draw=black, fill=black!50] (4.09, 1.57) rectangle (4.22, 3);
                    \fill[draw=red, fill=red] (3.59, 0.01) circle (0.0635);
                    \draw [<-] (3.5, 0.1) -- (3, 0.5) node[left] {\scriptsize Heater};
                    \draw [<-] (3.5, 1.55) -- (3, 1.55) node[left] {\scriptsize Load Cell};
                    \draw [<-] (3.75, 2.5) -- (3, 2.5) node[left] {\scriptsize Lowering Arm};
                    \draw[->]         (0.1, 0.6) -- (0.75, 0.6);
                    \draw[->]         (0.1, 1.2) -- (0.75, 1.2);
                    \node[right] at (-0.75, 1.5) {\scriptsize Inlet};
                    \draw[->]         (0.1, 1.8) -- (0.75, 1.8);
                    \draw[->]         (0.1, 2.4) -- (0.75, 2.4);
                    \draw[draw=black, dashed] (3.336, 0) rectangle (4.54, 0.34);
                    \draw[->]         (4.3, 1) -- (4.0, 0.34);
                    \node[right, align=left] at (4.3, 1) {\scriptsize Computational\\ \scriptsize Domain};
            \caption{Diagram of the experimental wind tunnel apparatus. Air flows through the wind tunnel from left to right. The dashed region represents the domain subset used for computational efforts.}
            \label{fig:speciesWindTunnel}

\section{Results and Discussion}
    \label{sec:results3}
    The flaming or non-flaming ignition outcome of each test and the logistic regression results for Douglas-fir wood, oak wood, and pine wood are shown in Figure~\ref{fig:logistic_plot} for the quiescent tests. The circular markers represent the outcome of each test, either flaming or no ignition. The logistic regression for each material is shown as a solid line with the shaded regions representing the 95\% confidence interval for each regression. Douglas-fir bark, pine bark, and wheat straw are not shown in Figure~\ref{fig:logistic_plot} because flaming ignition was not observed for five tests at the maximum heater temperature of 750\si{\celsius}. The heater temperature estimated to produce 50\% ignition probability from the logistic regressions for each material is shown in Figure~\ref{tab:composition50temp}. Anecdotally, self sustained smoldering was observed for all five materials at temperatures lower than the flaming ignition temperature. The heater temperature corresponding the onset of smoldering ignition was not measured in this study. 
        \begin{figure}[htpb]
            \centering
            \includegraphics[width=\figureWidthSet, trim={0 0 37cm 0}, clip]{conference_results_binary.png}
            \caption{Test results for materials where flaming ignition was observed. The circular markers denote individual tests. The solid lines represent the logistic regression and the shaded zones the 95\% confidence interval with the 50\% probability of ignition labelled}
            \label{fig:logistic_plot}
        \end{figure}
    Two aspects of the ignition results are of note from the quiescent cases. As anticipated, the differences in estimated heater temperature to produce 50\% ignition probability suggest that significant differences in ignition occur across a range of materials in the same apparatus and experimental conditions. Second, the transition between no ignition and flaming ignition may become less defined (i.e., the 95\% confidence interval of the regression spans a much larger temperature range) as the ignition temperature decreases. This is significant because it may cause more difficulty in predicting the ignition of materials that ignite at lower temperatures. Lower confidence in predicting ignition for easily ignitable materials would be detrimental to the usefulness of a model or predictive tool that may be implemented from this testing methodology. It is, however, unclear if the ignition to no ignition transition in other materials of similarly low ignition temperature will behave similarly to that of Douglas-fir. Characterization of additional materials is needed to elucidate these trends.

    Figure~\ref{fig:composition_plot} shows the averaged cellulose, hemicellulose, and lignin concentrations for each material according to the samples recorded in the Bioenergy Feedstock Library as shown in Table~\ref{tab:composition}. The marker colors for each material correspond to the calculated 50\% ignition probability for the quiescent tests except for the wheat straw and Douglas-fir bark which are denoted as 800\si{\celsius} to indicate that the 50\% ignition probability was higher than the temperatures tested.  Note that due to scaling inherent to a ternary projection the proportion of the fuels characterized as other causes a shift in the data. For example, the wheat straw and Douglas-fir are estimated to have the same cellulose content but do not lie on the same iso-line of 44\% cellulose. Nonetheless, the relative proportions of each constituent are retained and comparisons may still be made. 

@@

    Perhaps unsurprisingly, Douglas-fir bark and pine bark, with the highest lignin content, were not observed to ignite for the conditions tested. This suggests that materials with high lignin contents may be the least likely to ignite when exposed to embers and thus are the safest with respect to flaming ignition risk around homes. This does not, however, account for the potential for smoldering ignition and a subsequent smoldering to flaming transition that may occur. Thermogravametric analysis experiments have shown that while lignin begins to decompose at the lowest temperature of the three primary components, it decomposes at the slowest rate and at a much wider range of temperatures~\cite{Yang2007a}. It was also observed that the peak gas production of \ce{CO}, \ce{CO2}, and \ce{CH4} occurs at a higher temperatures than hemicellulose and cellulose. In contrast to Douglas-fir bark, wheat straw had the lowest estimated lignin concentration but was also not observed to ignite. When comparing wheat straw to Douglas-fir wood the higher lignin concentration suggests that Douglas-fir wood is likely to have a higher ignition temperature than wheat straw. This discrepancy is attributed to two material attributes. First, the pyrolysis process is complex and there is not a clear linear relationship between the composition of the fuel and flaming ignition. It appears that materials of different compositions are capable of producing gaseous products of near equal ignitability but the proportion of each constituent for which ignition to occurs most readily is unclear. Second, the thermal conductivity of the materials likely varies significantly which impacts the temperature gradients and mass of the material above the pyrolysis temperature, further obfuscating the effect of composition on ignition. 
    Fortunately, both thermal conductivity and chemical composition are measurable material properties and can be simultaneously considered when studying ignition.In summary, characterizing the flaming ignition propensity of fuels with respect to chemical composition alone will likely not yield satisfactory results. The inclusion of other material properties, such as thermal conductivity, however, may add enough additional knowledge for comprehensive predictions of ignition to be made. Further investigation of the differences in chemical and thermal properties of the materials is necessary to unravel the interactions between thermal and chemical processes and the resulting impact on flaming ignition before knowledge sufficient for a predictive tool that may be used for fire prevention and management may be created

@@

### 2022-04-08 manuscript4.tex
Updates to species paper results

\section{Abstract}
    some text
\section{Introduction}
    The combination of increasing wildfire severity and an increase of homes in the Wildland Urban Interface (WUI) has significantly increased the occurrence of home loss due to wildfires since the turn of the century~\cite{Manzello2013}. Ignition of fuels by firebrands that leads to the loss of a structure is a primary source of home loss~\cite{Koo2010a, Syphard2019, Roberts2021}, and in some fires has been the source of 2/3 of structure ignitions~\cite{Maranghides2011AFires}. Thus, it is important to understand the ignition of fuel beds by firebrands in order to mitigate the risk of ignition~\cite{Manzello2014}. Ideally, the understanding of the ignition process would result in a predictive model that would enable homeowners, firefighters, and other risk management personnel to determine location specific strategies for preventing ignition before and during a fire. Unfortunately, such a model does not currently exist due to critical shortfalls in the current knowledge of fuel bed ignition.

>>>>

\section{Abstract}
    some text
\section{Introduction}
    The combination of increasing wildfire severity and an increase of homes in the Wildland Urban Interface (WUI) has significantly increased the occurrence of home loss due to wildfires since the turn of the century~\cite{Manzello2013}. Ignition of fuels by firebrands that leads to the loss of a structure is a primary source of home loss~\cite{Koo2010a, Syphard2019, Roberts2021}, and in some fires has been the source of 2/3 of structure ignitions~\cite{Maranghides2013NISTIgnitions}. Thus, it is important to understand the ignition of fuel beds by firebrands in order to mitigate the risk of ignition~\cite{Manzello2014}. Ideally, the understanding of the ignition process would result in a predictive model that would enable homeowners, firefighters, and other risk management personnel to determine location specific strategies for preventing ignition before and during a fire. Unfortunately, such a model does not currently exist due to critical shortfalls in the current knowledge of fuel bed ignition.
